The Power of Two Random Choices\\ \large A follow up seminar to the Parallel Algorithms lecture\footnote{Jesper Larsson Tr\"aff and Francesco Versaci, lecture ``Parallel Algorithms'', 2012 winter term at TU Wien.}

### 2014-03-11 article/article.tex
enforce no line breaks before references and citations

\section{Introduction}
\label{sec:intro}
Suppose that $m$ balls are to be thrown into $n$ bins, according to some placement strategy. We are interested in the number of balls in the fullest bin, also called the \emph{maximum load}. Equivalently, one can look at the \emph{additive gap} of the allocation, i.e., the difference of the number of balls in the fullest bin and the average amount of balls in the bins. These balls-into-bins games (BiB) have been studied extensively (see, e.g., \cite{JK77}). In the context of Computer Science, the BiB model is useful to study, e.g., strategies for distributing tasks (balls) to processors (bins) as evenly as possible without having to query all the processors for the number of tasks currently assigned to them.

@@

\subsection{Single-choice BiB}
\label{sec:single-choice}
The \emph{Single-choice} strategy is a simple allocation scheme for the BiB problem and chooses one bin uniformly and independently at random for each ball. Using this strategy, the maximum load of the fullest bin will be
\begin{comment}
\footnote{A more accurate result exists: The maximum load is $\load = \Gamma^{-1}\left(n\right)\left(1+O\left(\frac{1}{\log \Gamma^{-1}\left(n\right)}\right)\right)$ \cite{G91}.}
\end{comment}
\begin{align*}
\load = \frac{\ln n}{\ln \ln n}  \left(1 + o\left(1\right)\right)
\end{align*}
for the case $m=n$, with high probability\footnote{We say that an event $\mathcal E_n$ occurs \emph{with high probability} (w.h.p.) if $\Pr\left[\mathcal E_n \right] \geq 1 - n^{-\alpha}$ for some arbitrary constant $\alpha > 0$.}~\cite{RS98}. This implies (see Theorem \ref{theorem:asymptotic} in Appendix \ref{sec:asymptotic})
\begin{align*}
\load\in \Theta\left(\frac{\ln n}{\ln \ln n}\right).
\end{align*}
Note that, since the average load is 1, $\delta \in \Theta\left( L_{max} \right)$ in this case. For the heavily loaded case, w.h.p.~the maximum gap is ~\cite{RS98}
\begin{align*}
\delta = \Theta\left(\sqrt{\frac{m  \ln n}{n}}\right).
\end{align*}

@@

Consider the slightly more involved \emph{Always-go-left} scheme: Suppose that the $n$ bins are partitioned evenly into $d$ groups, each of size $\lfloor \frac{n}{d}\rfloor$ or $\left\lceil\frac{n}{d}\right\rceil$ (To simplify the presentation, we will assume that $d$ divides $n$ evenly in the remainder of this survey). We select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. For tie-breaking, we assume a fixed ordering of bins $i$ for $1 \leq i \leq n$ and put the ball into the bin with smallest index $i$ among those with same minimal load. Somewhat surprisingly, V\"ocking \cite{VOC03} showed that such a strategy employing asymmetric tie-breaking in combination with non-uniform bin selection, reduces the bound for the gap to\
\begin{align*}
\delta = \frac{\ln\ln n}{d  \ln \Phi_d} + O(1),
\end{align*}
w.h.p., for all $m$. (The symbol $\Phi_d$ denotes a generalization of the golden ratio, based on generalized Fibonacci numbers. Refer to Appendix \ref{sec:fibonacci} for the definition.) Not only is this a significant reduction of the asymptotic bound, V\"ocking was able to show that the Always-go-left strategy is optimal up to additive constants \cite{VOC03}.

@@

\section{Classification}
\label{sec:classification}
Several different variations of this problem exist, which may be classified as follows:
\begin{compactitem}
\item In a \emph{sequential} process, the balls arrive one after another and ball is dealt with only after the previous one has been placed. Conversely, in a \emph{parallel} process, balls arrive in sets and the balls in one set are dealt with simultaneously \cite{ABS98}. 
\item In a \emph{static} problem, the balls are placed into the bins and no balls are ever removed, whereas in a \emph{dynamic} problem balls may also be removed from the system. In the random deletions model, for each newly inserted ball another one is chosen uniformly and independently at random and removed from the system \cite{ABKU99, MRS01}. More elaborate schemes exists: An oblivious adversary might specify the sequence of insertions and deletions of balls in advance, without any knowledge of their placement \cite{CFM+98}. Adler et al.~\cite{ABS98} studied another approach in the context of the parallel BiB problem, where the balls are stored in FIFO-order and at every time step the first ball in each of the bins is deleted. 
\item A \emph{finite} process is analyzed for a given time interval that is known beforehand, whereas for an \emph{infinite} one, placements and deletions are given for an infinite time span \cite{ABS98}.
\item In the lightly loaded case, $m$ is in the order of $n$, whereas in the heavily loaded case, $m$ is arbitrarily large. W.l.o.g.~we will assume that $m = n$ for the lightly loaded case. 
\end{compactitem}

@@

It is easy to see how the BiB model can be used: Assuming a perfect hash function\footnote{I.e., each element is mapped to one of the $n$ table entries uniformly and independently at random}, the elements can be seen as balls to be mapped to table entries (bins), implying a worst case look-up time of $\Theta\left(\frac{\ln n}{\log \ln n}\right)$, w.h.p., where $n$ denotes the number of elements. Applying the two-choice paradigm by using two hash functions and mapping elements to the least full of two table entries, results in a look-up time of $\Theta\left(\ln \ln n\right)$, w.h.p. This approach is easy to parallelize, it does not require re-hashing of data and requires only two hash functions \cite{ABKU99, MRS01}.

@@

To maximize the throughput of the servers, a uniform distribution of tasks to servers is desirable. While a central dispatcher can easily achieve this, it also poses a bottleneck within a distributed system, since querying $m$ servers for their current load involves significant overhead (e.g., sending a message to each server and waiting for the reply). As with hashing, the BiB model can be applied easily: we need to assign tasks (the balls) to servers (the bins). Assuming tasks are assigned to servers uniformly and independently at random, the two-choice paradigm is directly applicable: If each client samples two random servers for their load and sends its request to the least loaded server, the overhead is comparatively small and the difference in load on the $n$ servers is at most $O\left(\ln \ln n\right)$, w.h.p., \cite{KLM92,RS98, MRS01}.

\subsection{Emulation of shared memory machines on distributed memory machines}
\label{sec:DMM}
The Parallel Random Access Machine model (PRAM) ~\cite{P03} is a high level abstraction of a shared memory machine, in which memory access by any processor to any memory location is assumed to require constant time. It is extremely useful for the study of parallel algorithms. However, a PRAM is much harder to implement in hardware than a Distributed Memory Machine (DMM). Thus, ways to emulate a PRAM on a DMM have been studied extensively. Such an emulation distributes the processors and memory cells of the PRAM to the processors and memory modules (typically one per processor) of the DMM, with the goal to minimize the slowdown of the emulation, i.e., the time the DMM requires to simulate one step of the PRAM execution. Each memory module can handle only one memory access at a time step and more accesses to the same module will result in memory contention. The two-choice paradigm is useful for managing communication between the processors and the memory modules such that memory contention is minimal \cite{KLM92}.

@@

In this context, Valiant's paradigm \cite{V82} states that any such permutation routing problem can be solved by transforming it into two problems: First, find a path for each request to an intermediate destination chosen uniformly and independently at random. Next, find a path from the intermediate to the actual destination. This routing technique is analogous to the single-choice BiB problem where each communication request (or ball) chooses an intermediate destination (a bin) at random.
Extensive research \cite{CMM+98,MRS01} shows that by applying the two-choice paradigm, the link congestion can be bounded by $\Theta\left(\log \log n\right)$ , w.h.p.

@@

\section{Multiple-choice BiB}
\label{sec:multiple-choice}
The single-choice BiB strategy (Section \ref{sec:single-choice}) chooses one bin uniformly and independently at random for each ball. A multiple-choice BiB strategy generalizes this idea by  choosing $d \geq 2$ bins and placing the ball into the least full bin. We distinguish placement algorithms depending on how the $d$ locations are sampled \cite{VOC03}.

@@

\begin{theorem}
\label{theorem:greedy}
Consider any sequence of insertions and deletions s.t.~at most $n$ balls are in the bins at any time. If the balls are placed into bins by the Greedy algorithm, the maximum load is at most
\begin{align*}
\load = \frac{\ln \ln n}{\ln d}+ O(1)
\end{align*}
w.h.p.~for all $t$ \cite{ABKU99, BCSV06}.
\end{theorem}

This result is remarkable in several ways:
\begin{compactitem}
\item The apparently small change made for the Greedy algorithm compared to the single-choice BiB scheme results in an exponential decrease of the gap, even for $d=2$. 
\item Each additional choice ($d > 2$) decreases the gap by only a constant factor \cite{MRS01}.
\end{compactitem}

\subsection{The Always-go-left scheme}
\label{sec:AlwaysGoLeft}
The Greedy strategy presented in Section \ref{sec:greedy} assumes that $d$ bins are chosen uniformly and independently at random. V\"ocking \cite{VOC03} was the first to study the impact of non-uniform and dependent selections of the $d$ bins. Surprisingly, a multiple-choice BiB strategy with a non-uniform and independent selection of the $d$ bins, such as the Always-go-left algorithm introduced in \cite{VOC03}, achieves better \cite{BCSV06} load balancing than the Greedy scheme.

@@

\begin{theorem}
\label{theorem:aglm}
Consider any sequence of insertions and deletions s.t.~at most $n$ balls are in the bins at any time. If the balls are placed into bins by the Always-go-left algorithm, the maximum load is at most 
\begin{align*}
\load = \frac{\ln\ln n}{d  \ln \Phi_d} + O(1)
\end{align*}
w.h.p.~for all $t$ \cite{VOC03}.
\end{theorem}

\section{Analysis}
\label{sec:analysis}
In this section, we will prove Theorems \ref{theorem:greedy} and \ref{theorem:aglm} in detail, using witness trees to provide an upper bound for the event that a bin contains too many balls. Our analysis closely follows the proofs in \cite{VOC03}.

@@

We will prove the bounds for $m=n$ for both the Greedy and the Always-go-left scheme in Sections \ref{sec:analysis:greedy} and \ref{sec:analysis:alg}, respectively. However, we use a simplifying assumption (Assumption \ref{assumption:independence}) in those two sections to provide a more clear presentation of the proof technique. Section \ref{sec:analysis:nondistinctBalls} will extend these proofs to remove the simplifying assumption we make initially.

We will use a proof technique called witness trees, which allows for a clean analysis of both problems and is applicable to a wide variety of BiB scenarios. Other proof techniques such as layered induction and a differential equations approach have been used. See, e.g., \cite{MRS01} for an overview of different proof techniques for the BiB problem. The proof for arbitrarily large $m$ is more complicated and we refer the interested reader to \cite{BCSV06, TW13}.

@@

\subsubsection{Symmetric witness trees}
\label{sec:analysis:definitionSymWT}
A symmetric witness tree of order $L$ is a complete $d$-ary tree with $d^{L}$ leaves, where each node $v$ references a ball $b_v = \mathrm{ball}(v)$ and the nodes and edges of the witness tree represent events that may occur or not. In general, the same ball may be referenced by several nodes. However, initially we will make use of the following assumption in the analysis.
\begin{assumption}
\label{assumption:independence}
All events represented by the nodes and edges of a witness tree are stochastically independent. 
\end{assumption}
In other words, we assume that no ball is referenced by more than one node of a witness tree. A correct analysis will have to consider balls assigned to more than one node, but this simplifying assumption is useful to highlight the differences between the symmetric and asymmetric allocation schemes. Section \ref{sec:analysis:nondistinctBalls} will extend the analysis to witness trees with redundant balls.

@@

\begin{lemma}\label{lemma:swt:activation}
For some arbitrary $\alpha$ and large enough $L$, the activation of a symmetric witness tree of order $L$ occurs with probability at most $n^{-\alpha}$.
\end{lemma}
Speaking less formaly, Lemma \ref{lemma:swt:activation} states that for large enough $L$, the activation of a symmetric witness tree is unlikely.

@@

The number of different witness trees $w$ is given by the number of ways to assign balls to its nodes. The ball assigned to the root node can be chosen from at most $n$ balls, since by definition (Theorem \ref{theorem:greedy}), at most $n$ balls are in the bins at any time $t$ and a ball assigned to a node has to be in some bin at time $t$. By the same argument, each child node of the root is assigned one out of at most $n$ balls too. Applying this argument for each level of the witness tree, there are at most $w = n^k$ possibilities to assign balls to a witness tree, with $k = d^{L+1}-1$ being the number of nodes in a $d$-ary witness tree of order $L$.

An edge\footnote{where $v$ is the $i$-th child of $u$} $(u, v)$ is activated with probability at most $d/n$, which is the probability that the $i$-th location of the ball $b_u$ points to a particular location of the ball $b_v$ times the number of locations $d$. This implies a probability of at most
\begin{align*}
p_e = \left(\frac{d}{n}\right)^{k-1}
\end{align*} 
for the event that all $k-1$ edges of a witness tree are activated. Note that  the assumption that all balls are distinct (Assumption \ref{assumption:independence}) implies that all edge events are independent.

@@

Then, 
\begin{align*}
p_{swt} &\leq n  2^{{-d}^{\log_d\log_2 n + \log_d\left(1+\alpha\right)}} \\
		&= n  2^{-{d}^{\log_d\log_2 n} d^{\log_d \left(1+\alpha\right)}} \\
		&= n  \left( 2^{-\log_2 n  \left(1+ \alpha \right)} \right) \\
		&= n  n^{-\left(1+\alpha\right)} = n^{-\alpha} \neqn{swt:result}
\end{align*}
Thus, for any constant $\alpha$ and large enough $L$ as specified in Equation \ref{eqn:swt:chooseL} the probability that a symmetric witness tree of order $L$ is activated is bounded above by $p_{swt }\leq n^{-\alpha}$.
\end{proof}

Since the probabilty for the activation of a symmetric witness tree of order $L$ upper-bounds the probability of the bad event that some bin contains more than $L+3$ balls, Equation \ref{eqn:swt:result} is also an upper bound for the latter. Note that Equation \ref{eqn:swt:chooseL} can be written as
\begin{align*}
L &\geq  \frac{\ln\log_2 n}{\ln d} + \frac{\ln(1+\alpha)}{\ln d} \\
   &= \frac{\ln \log_2 n}{\ln d}  \left(1 + \frac{\ln\left(1+\alpha\right)}{\ln \log_2 n}\right) \\
  &=  \frac{\ln \ln n}{\ln d}  \left( 1+ o\left( 1\right)\right) \neqn{swt:boundL}
\end{align*}

This proves Theorem \ref{theorem:greedy} for the case $m=n$, under the assumption that all events are stochastically independent (Assumption \ref{assumption:independence}). \qed

\subsection{Always-go-left}
\label{sec:analysis:alg}
Similarly to Section \ref{sec:analysis:greedy}, where we used symmetric witness trees to upper-bound the probability that a bin contains too many balls when using the Greedy scheme, we will use asymmetric witness trees to get an analogous upper bound for the Always-go-left scheme. We will make the same simplification as stated in Assumption \ref{assumption:independence}, namely that all the events represented by the edges and nodes of a witness tree are stochastically independent.

@@

A $d$-ary asymmetric witness tree of order $L$ has the structure of the Fibonacci tree $T_d(d  L+1)$ and by Equation \ref{eqn:goldenFib} has $F_d(d L + 1) \geq \Phi_d^{d L-1}$ leaves.

@@

\begin{proof}
Analogously to the symmetric witness tree, the possible number of asymmetric witness trees is upper-bounded by $w=n^k$ and the probability that all edge events occur is\footnote{As in Section \ref{sec:analysis:greedy}, $k$ denotes the number of nodes of the witness tree and $l$ the number of leaves.} 
\begin{align*}
p_e = \left(\frac{d}{n}\right)^{k-1}
\end{align*}

@@

The above results imply the same probability for the activation of an asymmetric witness tree as we derived in Section \ref{sec:analysis:probabilitySymWT} for the activation of a symmetric witness tree, that is
\begin{align*}
p_{awt} &\leq n  2^{-l}
\end{align*}
However, the number of leaves is larger, since 
\begin{align*}
l \geq \Phi_d^{d L -1}
\end{align*}
holds\footnote{see Section \ref{sec:analysis:definitionAsymWT} for details}.

For any constant $\alpha > 0$, let 
\begin{align*}
L &\geq \left\lceil{\frac{\ln\log_2 n + \ln\left(1+\alpha\right)}{d \ln \Phi_d}}\right\rceil+\frac{1}{d} \neqn{awt:chooseL}
\end{align*}
\todo{the paper uses $+1$, instead of $+1/d$, but I can't workout the below result then} 
Then, 
\begin{align*}
p_{awt} &\leq n  2^{-\Phi_d^{d L -1}} \\
        &= n  2^{-\Phi_d^{d \left(\left\lceil{\frac{\ln\log_2 n + \ln\left(1+\alpha\right)}{d \ln \Phi_d}}\right\rceil+\frac{1}{d}\right) -1}} \\
        &\leq n  2^{-\Phi_d^{d \left({\frac{\ln\log_2 n + \ln\left(1+\alpha\right)}{d \ln \Phi_d}}+\frac{1}{d}\right) -1}} \\
        &= n  2^{-\Phi_d^{{\frac{\ln\log_2 n + \ln\left(1+\alpha\right)}{d \ln \Phi_d}}}} \\
        &= n  2^{-\Phi_d^{\log_{\Phi_d}\log_2 n + \log_{\Phi_d}\left(1+\alpha\right)}} \\
         &= n  2^{-\Phi_d^{\log_{\Phi_d}\log_2 n} \Phi_d^{ \log_{\Phi_d}\left(1+\alpha\right)}} \\
         &= n  2^{-\log_2 n \left(1+\alpha\right)} \\
         &= n  n ^{-(1+\alpha)} = n^{-\alpha} \neqn{awt:result}
\end{align*}
Thus, for any constant $\alpha$ and large enough $L$ as specified in Equation \ref{eqn:awt:chooseL} the probability that a symmetric witness tree of order $L$ is activated is bounded above by $p_{awt }\leq n^{-\alpha}$.
\end{proof}

>>>>

\section{Introduction}
\label{sec:intro}
Suppose that $m$ balls are to be thrown into $n$ bins, according to some placement strategy. We are interested in the number of balls in the fullest bin, also called the \emph{maximum load}. Equivalently, one can look at the \emph{additive gap} of the allocation, i.e., the difference of the number of balls in the fullest bin and the average amount of balls in the bins. These balls-into-bins games (BiB) have been studied extensively (see, e.g.,~\cite{JK77}). In the context of Computer Science, the BiB model is useful to study, e.g., strategies for distributing tasks (balls) to processors (bins) as evenly as possible without having to query all the processors for the number of tasks currently assigned to them.

@@

\subsection{Single-choice BiB}
\label{sec:single-choice}
The \emph{Single-choice} strategy is a simple allocation scheme for the BiB problem and chooses one bin uniformly and independently at random for each ball. Using this strategy, the maximum load of the fullest bin will be
\begin{comment}
\footnote{A more accurate result exists: The maximum load is $\load = \Gamma^{-1}\left(n\right)\left(1+O\left(\frac{1}{\log \Gamma^{-1}\left(n\right)}\right)\right)$~\cite{G91}.}
\end{comment}
\begin{align*}
\load = \frac{\ln n}{\ln \ln n}  \left(1 + o\left(1\right)\right)
\end{align*}
for the case $m=n$, with high probability\footnote{We say that an event $\mathcal E_n$ occurs \emph{with high probability} (w.h.p.) if $\Pr\left[\mathcal E_n \right] \geq 1 - n^{-\alpha}$ for some arbitrary constant $\alpha > 0$.}~\cite{RS98}. This implies (see Theorem \ref{theorem:asymptotic} in Appendix \ref{sec:asymptotic})
\begin{align*}
\load\in \Theta\left(\frac{\ln n}{\ln \ln n}\right).
\end{align*}
Note that, since the average load is 1, $\delta \in \Theta\left( L_{max} \right)$ in this case. For the heavily loaded case, w.h.p.~the maximum gap is ~\cite{RS98}
\begin{align*}
\delta = \Theta\left(\sqrt{\frac{m  \ln n}{n}}\right).
\end{align*}

@@

Consider the slightly more involved \emph{Always-go-left} scheme: Suppose that the $n$ bins are partitioned evenly into $d$ groups, each of size $\lfloor \frac{n}{d}\rfloor$ or $\left\lceil\frac{n}{d}\right\rceil$ (To simplify the presentation, we will assume that $d$ divides $n$ evenly in the remainder of this survey). We select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. For tie-breaking, we assume a fixed ordering of bins $i$ for $1 \leq i \leq n$ and put the ball into the bin with smallest index $i$ among those with same minimal load. Somewhat surprisingly, V\"ocking~\cite{VOC03} showed that such a strategy employing asymmetric tie-breaking in combination with non-uniform bin selection, reduces the bound for the gap to\
\begin{align*}
\delta = \frac{\ln\ln n}{d  \ln \Phi_d} + O(1),
\end{align*}
w.h.p., for all $m$. (The symbol $\Phi_d$ denotes a generalization of the golden ratio, based on generalized Fibonacci numbers. Refer to Appendix \ref{sec:fibonacci} for the definition.) Not only is this a significant reduction of the asymptotic bound, V\"ocking was able to show that the Always-go-left strategy is optimal up to additive constants~\cite{VOC03}.

@@

\section{Classification}
\label{sec:classification}
Several different variations of this problem exist, which may be classified as follows:
\begin{compactitem}
\item In a \emph{sequential} process, the balls arrive one after another and ball is dealt with only after the previous one has been placed. Conversely, in a \emph{parallel} process, balls arrive in sets and the balls in one set are dealt with simultaneously~\cite{ABS98}. 
\item In a \emph{static} problem, the balls are placed into the bins and no balls are ever removed, whereas in a \emph{dynamic} problem balls may also be removed from the system. In the random deletions model, for each newly inserted ball another one is chosen uniformly and independently at random and removed from the system~\cite{ABKU99, MRS01}. More elaborate schemes exists: An oblivious adversary might specify the sequence of insertions and deletions of balls in advance, without any knowledge of their placement~\cite{CFM+98}. Adler et al.~\cite{ABS98} studied another approach in the context of the parallel BiB problem, where the balls are stored in FIFO-order and at every time step the first ball in each of the bins is deleted. 
\item A \emph{finite} process is analyzed for a given time interval that is known beforehand, whereas for an \emph{infinite} one, placements and deletions are given for an infinite time span~\cite{ABS98}.
\item In the lightly loaded case, $m$ is in the order of $n$, whereas in the heavily loaded case, $m$ is arbitrarily large. W.l.o.g.~we will assume that $m = n$ for the lightly loaded case. 
\end{compactitem}

@@

It is easy to see how the BiB model can be used: Assuming a perfect hash function\footnote{I.e., each element is mapped to one of the $n$ table entries uniformly and independently at random}, the elements can be seen as balls to be mapped to table entries (bins), implying a worst case look-up time of $\Theta\left(\frac{\ln n}{\log \ln n}\right)$, w.h.p., where $n$ denotes the number of elements. Applying the two-choice paradigm by using two hash functions and mapping elements to the least full of two table entries, results in a look-up time of $\Theta\left(\ln \ln n\right)$, w.h.p. This approach is easy to parallelize, it does not require re-hashing of data and requires only two hash functions~\cite{ABKU99, MRS01}.

@@

To maximize the throughput of the servers, a uniform distribution of tasks to servers is desirable. While a central dispatcher can easily achieve this, it also poses a bottleneck within a distributed system, since querying $m$ servers for their current load involves significant overhead (e.g., sending a message to each server and waiting for the reply). As with hashing, the BiB model can be applied easily: we need to assign tasks (the balls) to servers (the bins). Assuming tasks are assigned to servers uniformly and independently at random, the two-choice paradigm is directly applicable: If each client samples two random servers for their load and sends its request to the least loaded server, the overhead is comparatively small and the difference in load on the $n$ servers is at most $O\left(\ln \ln n\right)$, w.h.p.,~\cite{KLM92,RS98, MRS01}.

\subsection{Emulation of shared memory machines on distributed memory machines}
\label{sec:DMM}
The Parallel Random Access Machine model (PRAM) ~\cite{P03} is a high level abstraction of a shared memory machine, in which memory access by any processor to any memory location is assumed to require constant time. It is extremely useful for the study of parallel algorithms. However, a PRAM is much harder to implement in hardware than a Distributed Memory Machine (DMM). Thus, ways to emulate a PRAM on a DMM have been studied extensively. Such an emulation distributes the processors and memory cells of the PRAM to the processors and memory modules (typically one per processor) of the DMM, with the goal to minimize the slowdown of the emulation, i.e., the time the DMM requires to simulate one step of the PRAM execution. Each memory module can handle only one memory access at a time step and more accesses to the same module will result in memory contention. The two-choice paradigm is useful for managing communication between the processors and the memory modules such that memory contention is minimal~\cite{KLM92}.

@@

In this context, Valiant's paradigm~\cite{V82} states that any such permutation routing problem can be solved by transforming it into two problems: First, find a path for each request to an intermediate destination chosen uniformly and independently at random. Next, find a path from the intermediate to the actual destination. This routing technique is analogous to the single-choice BiB problem where each communication request (or ball) chooses an intermediate destination (a bin) at random.
Extensive research~\cite{CMM+98,MRS01} shows that by applying the two-choice paradigm, the link congestion can be bounded by $\Theta\left(\log \log n\right)$ , w.h.p.

@@

\section{Multiple-choice BiB}
\label{sec:multiple-choice}
The single-choice BiB strategy (Section \ref{sec:single-choice}) chooses one bin uniformly and independently at random for each ball. A multiple-choice BiB strategy generalizes this idea by  choosing $d \geq 2$ bins and placing the ball into the least full bin. We distinguish placement algorithms depending on how the $d$ locations are sampled~\cite{VOC03}.

@@

\begin{theorem}
\label{theorem:greedy}
Consider any sequence of insertions and deletions s.t.~at most $n$ balls are in the bins at any time. If the balls are placed into bins by the Greedy algorithm, the maximum load is at most
\begin{align*}
\load = \frac{\ln \ln n}{\ln d}+ O(1)
\end{align*}
w.h.p.~for all $t$~\cite{ABKU99, BCSV06}.
\end{theorem}

This result is remarkable in several ways:
\begin{compactitem}
\item The apparently small change made for the Greedy algorithm compared to the single-choice BiB scheme results in an exponential decrease of the gap, even for $d=2$. 
\item Each additional choice ($d > 2$) decreases the gap by only a constant factor~\cite{MRS01}.
\end{compactitem}

\subsection{The Always-go-left scheme}
\label{sec:AlwaysGoLeft}
The Greedy strategy presented in Section~\ref{sec:greedy} assumes that $d$ bins are chosen uniformly and independently at random. V\"ocking~\cite{VOC03} was the first to study the impact of non-uniform and dependent selections of the $d$ bins. Surprisingly, a multiple-choice BiB strategy with a non-uniform and independent selection of the $d$ bins, such as the Always-go-left algorithm introduced in~\cite{VOC03}, achieves better~\cite{BCSV06} load balancing than the Greedy scheme.

@@

\begin{theorem}
\label{theorem:aglm}
Consider any sequence of insertions and deletions s.t.~at most $n$ balls are in the bins at any time. If the balls are placed into bins by the Always-go-left algorithm, the maximum load is at most 
\begin{align*}
\load = \frac{\ln\ln n}{d  \ln \Phi_d} + O(1)
\end{align*}
w.h.p.~for all $t$~\cite{VOC03}.
\end{theorem}

\section{Analysis}
\label{sec:analysis}
In this section, we will prove Theorems \ref{theorem:greedy} and \ref{theorem:aglm} in detail, using witness trees to provide an upper bound for the event that a bin contains too many balls. Our analysis closely follows the proofs in~\cite{VOC03}.

@@

We will prove the bounds for $m=n$ for both the Greedy and the Always-go-left scheme in Sections~\ref{sec:analysis:greedy} and~\ref{sec:analysis:alg}, respectively. However, we use a simplifying assumption (Assumption~\ref{assumption:independence}) in those two sections to provide a more clear presentation of the proof technique. Section~\ref{sec:analysis:nondistinctBalls} will extend these proofs to remove the simplifying assumption we make initially.

We will use a proof technique called witness trees, which allows for a clean analysis of both problems and is applicable to a wide variety of BiB scenarios. Other proof techniques such as layered induction and a differential equations approach have been used. See, e.g.,~\cite{MRS01} for an overview of different proof techniques for the BiB problem. The proof for arbitrarily large $m$ is more complicated and we refer the interested reader to~\cite{BCSV06, TW13}.

@@

\subsubsection{Symmetric witness trees}
\label{sec:analysis:definitionSymWT}
A symmetric witness tree of order $L$ is a complete $d$-ary tree with $d^{L}$ leaves, where each node $v$ references a ball $b_v = \mathrm{ball}(v)$ and the nodes and edges of the witness tree represent events that may occur or not. In general, the same ball may be referenced by several nodes. However, initially we will make use of the following assumption in the analysis.
\begin{assumption}
\label{assumption:independence}
All events represented by the nodes and edges of a witness tree are stochastically independent. 
\end{assumption}
In other words, we assume that no ball is referenced by more than one node of a witness tree. A correct analysis will have to consider balls assigned to more than one node, but this simplifying assumption is useful to highlight the differences between the symmetric and asymmetric allocation schemes. Section~\ref{sec:analysis:nondistinctBalls} will extend the analysis to witness trees with redundant balls.

@@

\begin{lemma}\label{lemma:swt:activation}
For some arbitrary $\alpha$ and large enough $L$, the activation of a symmetric witness tree of order $L$ occurs with probability at most $n^{-\alpha}$.
\end{lemma}
Speaking less formaly, Lemma~\ref{lemma:swt:activation} states that for large enough $L$, the activation of a symmetric witness tree is unlikely.

@@

The number of different witness trees $w$ is given by the number of ways to assign balls to its nodes. The ball assigned to the root node can be chosen from at most $n$ balls, since by definition (Theorem~\ref{theorem:greedy}), at most $n$ balls are in the bins at any time $t$ and a ball assigned to a node has to be in some bin at time $t$. By the same argument, each child node of the root is assigned one out of at most $n$ balls too. Applying this argument for each level of the witness tree, there are at most $w = n^k$ possibilities to assign balls to a witness tree, with $k = d^{L+1}-1$ being the number of nodes in a $d$-ary witness tree of order $L$.

An edge\footnote{where $v$ is the $i$-th child of $u$} $(u, v)$ is activated with probability at most $d/n$, which is the probability that the $i$-th location of the ball $b_u$ points to a particular location of the ball $b_v$ times the number of locations $d$. This implies a probability of at most
\begin{align*}
p_e = \left(\frac{d}{n}\right)^{k-1}
\end{align*} 
for the event that all $k-1$ edges of a witness tree are activated. Note that  the assumption that all balls are distinct (Assumption~\ref{assumption:independence}) implies that all edge events are independent.

@@

Then, 
\begin{align*}
p_{swt} &\leq n  2^{{-d}^{\log_d\log_2 n + \log_d\left(1+\alpha\right)}} \\
		&= n  2^{-{d}^{\log_d\log_2 n} d^{\log_d \left(1+\alpha\right)}} \\
		&= n  \left( 2^{-\log_2 n  \left(1+ \alpha \right)} \right) \\
		&= n  n^{-\left(1+\alpha\right)} = n^{-\alpha} \neqn{swt:result}
\end{align*}
Thus, for any constant $\alpha$ and large enough $L$ as specified in Equation~\ref{eqn:swt:chooseL} the probability that a symmetric witness tree of order $L$ is activated is bounded above by $p_{swt }\leq n^{-\alpha}$.
\end{proof}

Since the probabilty for the activation of a symmetric witness tree of order $L$ upper-bounds the probability of the bad event that some bin contains more than $L+3$ balls, Equation~\ref{eqn:swt:result} is also an upper bound for the latter. Note that Equation~\ref{eqn:swt:chooseL} can be written as
\begin{align*}
L &\geq  \frac{\ln\log_2 n}{\ln d} + \frac{\ln(1+\alpha)}{\ln d} \\
   &= \frac{\ln \log_2 n}{\ln d}  \left(1 + \frac{\ln\left(1+\alpha\right)}{\ln \log_2 n}\right) \\
  &=  \frac{\ln \ln n}{\ln d}  \left( 1+ o\left( 1\right)\right) \neqn{swt:boundL}
\end{align*}

This proves Theorem~\ref{theorem:greedy} for the case $m=n$, under the assumption that all events are stochastically independent (Assumption~\ref{assumption:independence}). \qed

\subsection{Always-go-left}
\label{sec:analysis:alg}
Similarly to Section~\ref{sec:analysis:greedy}, where we used symmetric witness trees to upper-bound the probability that a bin contains too many balls when using the Greedy scheme, we will use asymmetric witness trees to get an analogous upper bound for the Always-go-left scheme. We will make the same simplification as stated in Assumption~\ref{assumption:independence}, namely that all the events represented by the edges and nodes of a witness tree are stochastically independent.

@@

A $d$-ary asymmetric witness tree of order $L$ has the structure of the Fibonacci tree $T_d(d  L+1)$ and by Equation~\ref{eqn:goldenFib} has $F_d(d L + 1) \geq \Phi_d^{d L-1}$ leaves.

@@

\begin{proof}
Analogously to the symmetric witness tree, the possible number of asymmetric witness trees is upper-bounded by $w=n^k$ and the probability that all edge events occur is\footnote{As in Section~\ref{sec:analysis:greedy}, $k$ denotes the number of nodes of the witness tree and $l$ the number of leaves.} 
\begin{align*}
p_e = \left(\frac{d}{n}\right)^{k-1}
\end{align*}

@@

The above results imply the same probability for the activation of an asymmetric witness tree as we derived in Section~\ref{sec:analysis:probabilitySymWT} for the activation of a symmetric witness tree, that is
\begin{align*}
p_{awt} &\leq n  2^{-l}
\end{align*}
However, the number of leaves is larger, since 
\begin{align*}
l \geq \Phi_d^{d L -1}
\end{align*}
holds\footnote{see Section~\ref{sec:analysis:definitionAsymWT} for details}.

For any constant $\alpha > 0$, let 
\begin{align*}
L &\geq \left\lceil{\frac{\ln\log_2 n + \ln\left(1+\alpha\right)}{d \ln \Phi_d}}\right\rceil+\frac{1}{d} \neqn{awt:chooseL}
\end{align*}
\todo{the paper uses $+1$, instead of $+1/d$, but I can't workout the below result then} 
Then, 
\begin{align*}
p_{awt} &\leq n  2^{-\Phi_d^{d L -1}} \\
        &= n  2^{-\Phi_d^{d \left(\left\lceil{\frac{\ln\log_2 n + \ln\left(1+\alpha\right)}{d \ln \Phi_d}}\right\rceil+\frac{1}{d}\right) -1}} \\
        &\leq n  2^{-\Phi_d^{d \left({\frac{\ln\log_2 n + \ln\left(1+\alpha\right)}{d \ln \Phi_d}}+\frac{1}{d}\right) -1}} \\
        &= n  2^{-\Phi_d^{{\frac{\ln\log_2 n + \ln\left(1+\alpha\right)}{d \ln \Phi_d}}}} \\
        &= n  2^{-\Phi_d^{\log_{\Phi_d}\log_2 n + \log_{\Phi_d}\left(1+\alpha\right)}} \\
         &= n  2^{-\Phi_d^{\log_{\Phi_d}\log_2 n} \Phi_d^{ \log_{\Phi_d}\left(1+\alpha\right)}} \\
         &= n  2^{-\log_2 n \left(1+\alpha\right)} \\
         &= n  n ^{-(1+\alpha)} = n^{-\alpha} \neqn{awt:result}
\end{align*}
Thus, for any constant $\alpha$ and large enough $L$ as specified in Equation~\ref{eqn:awt:chooseL} the probability that a symmetric witness tree of order $L$ is activated is bounded above by $p_{awt }\leq n^{-\alpha}$.
\end{proof}
